---
title: "R Training 3"
output: html_notebook
---

# Exercise 9: Sperate the syllables in a song

Lets load our data. Remember to load any necessary libraries before you begin. 

```{r}
library(seewave)
zf_data = readWave('zfinch.wav')
tim = seq(1/zfFs, length(zf_data@left)/zfFs, 1/zfFs)
```

We are going to separate the syllables by amplitude. The problem with the original signal is that it goes both up and down, so it is not possible to directly measure the amplitude. Below I describe two methods,"Strategy 1" and "Strategy 2", that can be used to make a useful measure of the amplitude.

These two methods produce nearly the same result. The first method is to rectify the signal and low-pass filter, and the other is to use a function known as "Hilbert" and low-pass filter.

Both require that the signal be centered at zero. It is easy to do this by subtracting the mean from the original signal.

```{r}
zf_data_center = zf_data@left - mean(zf_data@left)
```

## Strategy 1: Rectify and smooth

First lets take the absolute value using **abs()**.

```{r}
rz = abs(zf_data_center)
```

Now we are going to smooth this signal using a median filter. R does this with the function **runmed()**. 

Before we do that lets set our variables. Here **sms** is the duration, in seconds, of the filter - I have set the default to 20 milliseconds. If **sms** is too long, then it will smooth out the signal too much. If **sms** is too short, then you will get spurious syllables. We will multiply sms with the sample rate, **Fs**. This will give us the number of samples for the duration that **sms** specifies.

```{r}
sms = 0.020
Fs = zf_data@samp.rate
```

The we apply our filter...

```{r}
mrz = runmed(rz, Fs*sms)
```

We don't need to do the next step, but it makes life a little easier later. Here we multiple by 1000 so that the values are ~ 1 rather than 0.001.

```{r}
mrz = mrz*1000
```

## Strategy 2: Hilbert and Smooth

This is almost identical to Strategy 1, but uses a nice function called **hilbert**. This function is from the package **seewave**, which you should already have installed!

```{r}
hz = abs(hilbert(rz, f = Fs))
```

Lets apply the same filter as above.

```{r}
mhz = runmed(hz, Fs*sms)
mhz = mhz*1000
```

Time to plot. First the original signal...

```{r}
plot(tim, zf_data@left, 
     main = "Zebra Finch Data", 
     col = "Blue", type = "l", 
     xlab = "Time" , 
     ylab = "", 
     yaxt = "none", 
     ylim = c(0,250))
axis(2, seq(27, 227,100))
```

...and the same signal after filtering.

```{r}
plot(tim, mhz, 
     type = "l", 
     col = "Green", 
     yaxt = "n", 
     ylab = "", 
     xaxt = "n", 
     xlab = "")
par(new = TRUE)
plot(tim, mrz, 
     type = "l", 
     col = "Red", 
     ylab = "", 
     xlab = "Time")
```


## Sepperate the Song into Syllables 


Animals often produce signals that are composed of multiple parts. One strategy to understand the signal is to divide it into syllables. That is what we will do here - use an amplitude threshold to separate each part into distinct syllables. 

We start by replotting the same data (**mrz**) from the last system, but the **y axis** will be on a log scale. This is acheived using the parameter **log = "y"** which plots with the X axis normal and Y axis on a log scale. You can make the same change to the x-axis with **log = "x"**. 

```{r}

plot(tim, mrz, 
     type = "l", 
     log = "y", 
     col = "Red", 
     ylab = "", 
     xlab = "Time")

```

Now that we've plotted our data, need to choose a threshold. We can either look at the plot and guess, or we can click on the plot and the computer will give us the coordinates where we clicked:

### First strategy - I guess "900"!

```{r}
thresh = 900
```

For clicking, we use the command **locator()** and specify how many clicks we are going to make, which is 1 in this case:

```{r}
thresh = locator(1)

```
 
If we used **locator()**, then **thresh** will have two values, the x-value and the y-value of the click. We only want the y-value, which is in the second position, and hence we only want
**thresh[2]**. We can do the following:

```{r}
thresh = thresh[2]
```

Let's replot with a line at the guessed threshold just to check. You might have to redo this section over and over again with different values for **thresh** until we get a value that we think will suffice.

**y** will be the amplitude of the threshold

```{r}
y = thresh
```

And now we plot the line with the **abline()** command using **h = ** to indicate a hoizontal line:

```{r}
plot(tim, mrz, type = "l", log = "y", col = "Red", ylab = "", xlab = "Time")
par(new = TRUE)
abline(h = y, col = "Blue")
```

Now we use the **which()** command to get the part of the signal above the threshold. The variable **syls** will get the index numbers for each value of **mrz** that is larger than **thresh**:

```{r}
syls = which(mrz > thresh)
```

Now we're going to do something that will extract the starts and ends of each syllable. This seems complicated, but it is simple.

The first step is to make a list of zeros that is the length of the signal. We can use the **rep()** command:

```{r}
zz = rep(0, length((zf_data@left)))
```

Let's look at **zz**:

Replot the original signal and then plot **zz**.

```{r}
par(mfrow = c(2,1))
plot(tim, zf_data@left, 
     main = "Zebra Finch Data", 
     col = "Blue", 
     type = "l", 
     xlab = "Time" , 
     ylab = "", 
     yaxt = "none", 
     ylim = c(0,250))
axis(2, seq(27, 227,100))

plot(tim, zz, 
     col = "Red", 
     type = "l", 
     xlab = "Time" , 
     ylab = "", 
     yaxt = "none")

```

From above, **syls** is the list of values that are above the threshold - these are where the syllables are located. We will set **zz** to 1 for each of these values...

```{r}
zz[syls] = 1
```

Now lets replot **zz**:

```{r}
par(mfrow = c(2,1))
plot(tim, zf_data@left, 
     main = "Zebra Finch Data", 
     col = "Blue", 
     type = "l", 
     xlab = "Time" , 
     ylab = "", 
     yaxt = "none", 
     ylim = c(0,250))
axis(2, seq(27, 227,100))

plot(tim, zz, 
     col = "Red", 
     type = "l", 
     xlab = "Time" , 
     ylab = "", 
     yaxt = "none")
```


Cool - these blocks roughly align with the syllables... but what we need are the start times and end times of the syllables. We will use a trick to get them...

**diff** takes the difference between adjacent values...

*For example sample(2) - sample(1)...*

```{r}
yy = diff(zz)
```

Now let's plot this. Starts are marked by **1** and ends are marked by **-1**.

```{r}
plot(tim[1:length(tim)-1], yy, 
     type = "l", 
     col = "Blue")
```
 

Now we use the **which()** command to get the *starts* and the *ends* of each syllable. Notice that we use **==** to indicate when we are **ASKING** if values are equal, but use **=** to **SET** a variable to a value.

```{r}
starts = which(yy == 1)
ends = which(yy == -1)
```


Now this is simple! We use a loop - the **for** command to get each syllable. Before we write our loop we will create some empty lists to hold our outputs.

*Note: Loops are generally not recommended in R, but here it is fairly simple and makes ourlives easier.*

```{r}
syllable = c()
timmy = c()
timm = c()


for (i in seq(1,length(starts))) {

  syllable[[i]] = zf_data@left[starts[i]:ends[i]]
  timmy[[i]] = tim[starts[i]:ends[i]]
  timm[[i]] = seq(1/Fs, (1 + ends[i] - starts[i])/Fs, 1/Fs)
}
```

**syllable[1]** will have the entire 1st syllable, **syllable[2]** will have the entire 2nd syllable, etc.

*Note: timmy and timm are not necessary, but might be useful.
**timmy** is the real times from the song.
**timm** is the time of the syllable from 0*

Just to show what has happened, here is a plot:

```{r}

par(mfrow = c(2,4))
spectro(syllable[[4]], Fs, wl = 512, palette = reverse.heat.colors, collevels = seq(-100, -10, 1), scale = FALSE)
spectro(syllable[[5]], Fs, wl = 512, palette = reverse.heat.colors, collevels = seq(-100, -10, 1), scale = FALSE)
spectro(syllable[[6]], Fs, wl = 512, palette = reverse.heat.colors, collevels = seq(-100, -10, 1), scale = FALSE)
spectro(syllable[[7]], Fs, wl = 512, palette = reverse.heat.colors, collevels = seq(-100, -10, 1), scale = FALSE)

plot(timmy[[4]], syllable[[4]], type = "l", col = "Red")
plot(timmy[[5]], syllable[[5]], type = "l", col = "Red")
plot(timmy[[6]], syllable[[6]], type = "l", col = "Red")
plot(timmy[[7]], syllable[[7]], type = "l", col = "Red")
```

I picked syllables 4,5,6, and 7. You can pick others if you wish.


Now we will get the silent parts between syllables. This can be important because sometimes the time between signals is an independent signal. For example, in frogs the duration between calls determines the "pulse repetition rate", which can indicate whether the call is a mate attraction signal or an aggressive signal.

The procedure is almost the same as for the syllables as above. What is different is that instead of copying the signal from **zfData**, we instead make pure silence by putting in a flat signal with an amplitude of 0.

We use a loop for each of the ends (which is the start of each silence).

To do this, make a list of zeros with the length of the interval and get the length of the interval by getting the difference between the end of the silence, which is the 'start' of the next syllable, and the start of the silence, which is the 'end' of the previous syllable. Complicated!?! NOt really!

```{r}

for (j in seq(2,length(ends))) {
  
  nop[[j]] = list(rep(0, times = starts[j] - ends[j-1]))
  noptimmy = tim[ends[j-1]:starts[j]]
  
}
```



###For review: You can easily get the data for each syllable.

**syllable[[1]]** is the data for the first syllable, and **syllable[[2]]** the second.

Let's plot a syllable.

```{r}

par(mfrow = c(1,2))
spectro(syllable[[7]], Fs, wl = 512, palette = reverse.heat.colors, collevels = seq(-100, -20, 1), scale = FALSE)
plot(timmy[[7]], syllable[[7]], type = "l", col = "Red")
```

 

How many syllables did we find??

**length(syllable)**

How long (duration) is syllable{7}?

**length(syllable[[7]])/Fs**

# Exercise 10: Make a NEW song with the syllable order reversed












