---
title: "R Training 2"
output:
  html_document:
    df_print: paged
---
# Introduction

Welcome back! In this tutorial we will learn how to apply filters to our signals, as wells as learn how to sort though data. 

```{r, include=FALSE}
library(signal)
library(tuneR)
library(ggplot2)
library(oce)
library(seewave)
library(gridExtra)
library(DFWE)
```



# Exercise 5: Filter a signal 

First let's filter an artificial signal. We'll add two sinewaves together:

one with a frequency of 5 kHz and the other at 500 Hz. We've done this before, so no problem!

```{r}
freq1 = 5000;
freq2 = 500;

sampleFs = 20000;
intvl = 1/sampleFs;
secs = 2;
tim = seq(intvl, secs, intvl)
amp1 = 1;
amp2 = 1;
wave1 = sin(tim*2*pi*freq1) * amp1;
wave2 = sin(tim*2*pi*freq2) * amp2;
addwav = wave1 + wave2;
```


We'll do a high-pass and a low-pass filter. We can have different frequencies for the high- and low-pass parts, but to start we'll use the same frequency, 2500Hz.


```{r}
lowcut = 2500
highcut = 2500
```


These next steps make the filter. There are many types of filter -- the one we are using is called a "Butterworth" filter. The only thing to change here is the order, **n** -- which is the slope of the filter. Lower numbers have a broader slope, whereas higher numbers have a steeper slope.

### Syntax 

We will be using the function: 

bwfilter(wave, f, n = 1, from = NULL, to = NULL,
bandpass = TRUE) 

We can specify a few parameters: **wave** specifies our wave file, **f** is the sampling frequency, **n** is the order of the filter, **from** and**to** give the start and end frequencies, respectively. Don't worry about **bandpass** right now.


```{r}
lowfilt = bwfilter(addwav, sampleFs, n = 5, from = NULL, to = 2500 )
highfilt = bwfilter(addwav, sampleFs, n = 5, from = 2500, to = NULL )
```


Now let's plot the results. 

```{r}

specplot(addwav, Fs = sampleFs) 

```

```{r}
specplot(lowfilt, Fs = sampleFs)

```

```{r}
specplot(highfilt, Fs = sampleFs)
```

# Exercise 6: Filter a signal - noise


The first step is to make a noisy signal. 
Set the length, in seconds

```{r}
len = 1
```

Set the SampleRate, in Hz. and make a time series...

```{r}
sampleFs = 20000

```


Make the noise sequence using **runif**. This function generates a series of random numbers within a set min and max value. 

### Syntax 

runif(number_of_numbers, min, max). If you want a random sequence taken from a normal distribution you can use rnorm. 

```{r}

noisy = runif(length(tim), min = -2^16, max = 2^16)

```


This is great - we can plot this

```{r}
specplot(noisy, Fs = sampleFs)
```


Now we filter, as above, this time using differnt frequencies for our filters. 

Our low pass filter will have a max frequency of 1500.
Our high pass filter will have a minimum frequency of 2700. 

```{r}
lowfilt_noise = bwfilter(noisy, sampleFs, n = 5, from = NULL, to = 1500 )
highfilt_noise = bwfilter(noisy, sampleFs, n = 5, from = 2700, to = NULL )
```

Now we can plot these as sonograms.

```{r}
specplot(lowfilt_noise, Fs = sampleFs) 
```

```{r}
specplot(highfilt_noise, Fs = sampleFs) 
```

### Exercise 6.1 

Vary the 'order' of the filter between 1 and 9. What happens?

Can you make a bandpass signal, where you filter out signal below 1000 Hz and above 7700 Hz??
